---
title: "Script_plots_18s"
author: "Charlotte Berthelier"
date: "24/05/2023"
output: html_document
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/shared/projects/glomicon/finalresults/finalresults_to_send/")
```

# Final table

```{r}
library(data.table)
library(magrittr)
library(dplyr)
library(ggplot2)
```

## Inport and merge ASV and taxo tables
# import taxo and parse the taxonomy item
```{r}
taxo_BH <- fread("outputs/asv_table/18s_dada2_v1.0.filtered.table.with.taxo.vsearch_BH.tsv.gz") # best-hit
taxo_idtaxa <- fread("outputs/asv_table/18s_dada2_v1.0.filtered.table.with.taxo.idtaxa.tsv.gz") # idtaxa

head(taxo_BH)
head(taxo_idtaxa)
colnames(taxo_idtaxa)
colnames(taxo_BH)

taxo_BH <- taxo_BH[,.SD,.SDcols=c("amplicon","taxonomy","identity","sequence",
                                   "total","spread",
                                   sort(grep("^ROS",colnames(taxo_BH),value=T)))]

setnames(taxo_BH,c("amplicon","taxonomy","identity"),c("BH_amplicon","BH_taxonomy","BH_identity"))
setnames(taxo_idtaxa,c("amplicon","taxonomy","sequence"),c("IdTaxa_amplicon","IdTaxa_taxonomy","IdTaxa_sequence"))
setnames(taxo_idtaxa,c("total","spread","confidence"),c("IdTaxa_total","IdTaxa_spread","IdTaxa_identity"))
```

# Split taxonomic best hit ranks (column "taxonomy")
## columns names : kingdom domain	phylum class	order	family genus species
```{r}
seq_assignement <- data.frame(Representative_sequence=taxo_BH$BH_taxonomy,
                              kingdom = sapply(strsplit(taxo_BH$BH_taxonomy, split="\\:"), "[", 1),
                              domain = sapply(strsplit(taxo_BH$BH_taxonomy, split="\\:"), "[", 2),
                              phylum = sapply(strsplit(taxo_BH$BH_taxonomy, split="\\:"), "[", 3),
                              class = sapply(strsplit(taxo_BH$BH_taxonomy, split="\\:"), "[", 4),
                              order = sapply(strsplit(taxo_BH$BH_taxonomy, split="\\:"), "[", 5),
                              family = sapply(strsplit(taxo_BH$BH_taxonomy, split="\\:"), "[", 6),
                              genus = sapply(strsplit(taxo_BH$BH_taxonomy, split="\\:"), "[", 7),
                              species = sapply(strsplit(taxo_BH$BH_taxonomy, split="\\:"), "[", 8),
                              stringsAsFactors = F)
seq_assignement2 = apply(seq_assignement, FUN=gsub, MARGIN=2, pattern = ",[a-z]", replacement = "") # remove letter after ","
seq_assignement2 = apply(seq_assignement2, FUN=gsub, MARGIN=2, pattern = "\\*", replacement = "NA") # empty field ("*") to NA

taxo_split_BH = cbind(taxo_BH,seq_assignement2[,-1]) # to eliminate created column named 'Representative_sequence'
head(taxo_split_BH)


seq_assignement_IdTaxa <- data.frame(Representative_sequence=taxo_idtaxa$IdTaxa_taxonomy,
                              kingdom = sapply(strsplit(taxo_idtaxa$IdTaxa_taxonomy, split="\\|"), "[", 2),
                              domain = sapply(strsplit(taxo_idtaxa$IdTaxa_taxonomy, split="\\|"), "[", 3),
                              phylum = sapply(strsplit(taxo_idtaxa$IdTaxa_taxonomy, split="\\|"), "[", 4),
                              class = sapply(strsplit(taxo_idtaxa$IdTaxa_taxonomy, split="\\|"), "[", 5),
                              order = sapply(strsplit(taxo_idtaxa$IdTaxa_taxonomy, split="\\|"), "[", 6),
                              family = sapply(strsplit(taxo_idtaxa$IdTaxa_taxonomy, split="\\|"), "[", 7),
                              genus = sapply(strsplit(taxo_idtaxa$IdTaxa_taxonomy, split="\\|"), "[", 8),
                              species = sapply(strsplit(taxo_idtaxa$IdTaxa_taxonomy, split="\\|"), "[", 9),
                              stringsAsFactors = F)

taxo_split_idtaxa = cbind(taxo_idtaxa,seq_assignement_IdTaxa[,-1]) # to eliminate created column named 'Representative_sequence'
head(taxo_split_idtaxa)

```

# Selection based on Sequence Identity, keep all ASVs if identity > 80
```{r}
taxo_BH <- taxo_split_BH[taxo_split_BH$BH_identity > 80,]
taxo_idtaxa <- taxo_split_idtaxa[taxo_split_idtaxa$IdTaxa_identity > 80,]

```


# generation of the sample list (usefull for graphs)
```{r}
samples <- data.table(sample = grep("^ROS",colnames(taxo_BH),value=T))
head(samples)
samples[,nreads:=sapply(sample,function(X) taxo_BH[,sum(get(X))])] # nb of reads per sample (= station_depth_fraction = sum column)
head(samples)
```

# Selection of asv present in > 1 samples (column spread) and with an occurence >1 (column total)
```{r}
taxo_BH_filtered <- taxo_BH[total>1&spread>1][order(total,decreasing = T)] # sort by alphabetical order

# Ranks: domain	phylum	class	order	family	genus	species	subcluster	clade	subclade
taxo_final_full_filtered <- taxo_BH_filtered[,.SD,.SDcols=c("BH_amplicon","sequence","BH_taxonomy","domain", "phylum", "class", "order", "genus", "BH_identity","total","spread",sort(grep("^ROS",colnames(taxo_BH_filtered),value=T)))]

```

# Figures
## Input
```{r}
asv_table <- taxo_final_full_filtered
```


## Taxo and palette
# Most abundant class and order
```{r}
# Select class (4) and order (5) for each sample
taxo_class <- asv_table[,tstrsplit(BH_taxonomy,split=",.:",keep = c(4,5))] # c(4,5)
taxoplot <- data.table(asv_table[,.(BH_amplicon,BH_identity,total)],taxo_class)
setnames(taxoplot,c("V1","V2"),c("class","order"))
taxoplot[order=="*",order:=paste(class,"X",sep="_")]

# selection of 20 most abundant orders
tmp <- taxoplot[,order] %>% table() %>% sort(decreasing = T)
tmp <- tmp[1:20]

taxo_palette <- taxoplot[order%in%names(tmp),list(class,order)] |> unique()
taxo_palette[,unique(class)]

# class palette
x <- RColorBrewer::brewer.pal(taxo_palette[,length(unique(class))],"Paired") # Set1, Set3
names(x) <- taxo_palette[,unique(class)]

# stretch the palette
taxo_palette[,class_col:=x[class]]

taxo_palette[,order_col:={colfunc <- colorRampPalette(c("white",unique(class_col),"black"))
               tmp <- colfunc(.N+4) # Generate palette de .N=nombre d'ordre + 4
               tmp[c(-1:-2,-(length(tmp)-1):-length(tmp))]},by=class] # delete first and last two

colman <- taxo_palette[,order_col]
names(colman) <- taxo_palette[,order]
colman <- c(colman,Others="grey90")

taxoplot[,orderplot:=order]
taxoplot[!order%in%names(colman),orderplot:="Others"]
x <- taxoplot[orderplot!="Others"][order(order)][order(class)][,unique(orderplot)]

# set colour order for legend and figure
taxoplot[,orderplot:=factor(orderplot,levels=c("Others",x))]
colman <- c(colman[x],colman["Others"])
```



```{r, fig.width=8,fig.align="center",fig.height=4,warning=F}
p1 <- ggplot(taxoplot,aes(x=BH_identity))+
  geom_bar(aes(fill=orderplot))+
  scale_fill_manual(values = colman)+
  scale_x_binned("% identity")+
  scale_y_continuous("# ASVs")+
  theme_classic()

p1.2 <- ggplot(taxoplot,aes(y=BH_identity))+
  geom_bar(aes(fill=orderplot))+
  scale_fill_manual(values = colman)+
  scale_x_continuous("ASV Nb")+
  scale_y_binned("% Identity")+
  theme(axis.text.y = element_text(size=7))+
  guides(fill=guide_legend(title="order"))

p2 <- ggplot(taxoplot,aes(x=BH_identity))+
  geom_bar(aes(fill=orderplot,weight=total))+
  scale_fill_manual(values = colman)+
  scale_x_binned("% identity")+
  scale_y_continuous("# reads")+
  theme_classic()

p2.2 <- ggplot(taxoplot,aes(y=BH_identity))+
  geom_bar(aes(fill=orderplot,weight=total))+
  scale_fill_manual(values = colman)+
  scale_x_continuous("Read nb")+
  scale_y_binned("% Identity")+
  theme(axis.text.y = element_text(size=7))+
  guides(fill=guide_legend(title="order"))

p1
# p1.2
p2
# p2.2

#pdf("outputs/graphes/diversity_barplots_18S.pdf",width = 8,height = 5)
#p1+p2+plot_layout(guides = "collect")
#dev.off()
```

```{r}
head(asv_table)

asv_table_long <- melt(asv_table,measure.vars = samples[,sample],id.vars = "BH_amplicon")

asv_table_long <- asv_table_long[value>0]
asv_table_long <- merge(asv_table_long,taxoplot[,list(BH_amplicon,orderplot)],by="BH_amplicon")
asv_table_long[,sampleplot:=paste(variable)] # a garder si on veut changer les noms des stations

asv_table_long_plot <- asv_table_long[,.(nreads=sum(value),nasvs=length(unique(BH_amplicon))),
                                       by=list(variable,orderplot,sampleplot)]

colnames(asv_table_long_plot)[1] <- "Samples"

# Export table
fwrite(asv_table_long_plot,"asv_table_long_plot.txt",sep="\t",na="NA")


p3 <- ggplot(asv_table_long_plot,aes(x=nreads,y=Samples))+
  geom_col(aes(fill=orderplot),position = "fill")+
  scale_fill_manual(values = colman,na.value="black")+
  scale_x_continuous("Percent Total Reads")+
  theme(axis.text.y = element_text(size=8))+
  guides(fill=guide_legend(title="Most abundant class"))

# tout avoir dans la même matrice
# colonne en + pour cutting
# faced grid

p3

```


```{r}
library(stats)
library(vegan)
library("FactoMineR")

df <- asv_table[,.SD,.SDcols=c(sort(grep("^ROS",colnames(asv_table),value=T)),"BH_amplicon")]
df <- data.frame(df, row.names = "BH_amplicon")
df <- t(df)

# abondance relative
d <- decostand(df, method = "total")
# Distance matrix
d2 <- vegdist(d, method="bray", binary=FALSE, diag=FALSE, upper=FALSE, na.rm = FALSE)

# Hierarchical clustering
hc <- hclust(d2, method ="ward.D2")

# Dendrogram
plot(as.dendrogram(hc))

vegdist(df, method="bray", binary=FALSE, diag=FALSE, upper=FALSE, na.rm = FALSE)
```